BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-03-23 Paper/D5.5/D5.5.tex
Small fix.
V 0.2 sent to internal reviewers.
Added Olivier's corrections.

\bibliographystyle{plain}
\begin{document}
\devnum{5.5}
\title{BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications}
\version{0.1}
\deadline{6}
\level{RE}
\nature{Report}
\coordinator{ (CNRS)}
\contributors{Thierry Deutsch, Luigi Genovese, Jean-François Méhaut, Kevin Pouget, Brice Videau}
\reviewers{None}
\keywords{BOAST, Deliverable}

@@

\begin{changelog}
\change{0.1}{Initial version of the deliverable}
\end{changelog}

@@

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in FORTRAN for historical and
performance reason. Codes can be quite huge (several thousands lines of code)
with lots of functionalities. Nonetheless they are usually based on computing
kernels.  Computing kernels are resource intensive and well defined parts of a
program that usually work on precisely defined data. Those kernels are the time
consuming part of an HPC application. They are thus the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes some of
those developers only work a few month on the application. Maintaining optimized
code written by someone else is quite a challenge. Several languages and
programming paradigms can also be used in a project and thus the maintainer must
be knowledgeable in several area of expertise. Portability problems can also be
cause by the availability of an optimizing compiler for a specific language and
architecture. C compilers are usually the first available while FORTRAN may
sometimes arrive later.

In Section~\ref{use_cases} we will present two HPC application that we used
as use cases: SPECFEM3D and BigDFT. They are both based on computing
kernels and were selected as candidate applications in the Mont-Blanc project.

@@

The race to exascale is not going to simplify the environment. All of the above
architectures can be considered. Network architectures are also part of the
architecture and can be very diverse.  For instance European FP7 project DEEP
considers using Accelerators (XEON Phi) while the European FP7 project
Mont-Blanc considers using low-power embedded processor with integrated GPU.

@@

This is a good example of an algorithm that is simple in its formulation but can
be complex to optimize because many different optimizations are available and
can be combined together. The correct combination of optimization will be
dependant on the targeted architecture.

@@

Listing~\ref{lst:laplace_c_filter} show a simple implementation of the Laplace
filter in C99. For clarity, boundary conditions have been omitted.

@@

\paragraph{Vectorization}
The first proposed optimization is to compute 5 pixels instead of one using the
vectors available to the Mali architecture. To be efficient the vectors have to
be directly loaded from memory. Using vectors of 16 elements 15 useful component
are simultaneously loaded and can be used in computation. This optimization
yields speedups between 1.5 and 6 depending on the image size.

@@

\paragraph{More Optimizations} Several other optimizations can be attempted at
this point. For instance reducing or increasing the number of pixel each work
item is processing. They can yield improved performance in some cases, especially
when avoiding memory alignment problems. These optimizations will not be shown
here but can be found in~\cite{opencl_arm_training}.

@@

Nonetheless, the process is frustrating. The optimizations are never evaluated
independently from one another. Some were arbitrarily configured (the number of
pixel chosen for instance). Testing the different combination of those
optimizations and the different parameters would prove very costly in developer
time and like all repetitive and tedious tasks, error prone. Thus every created
kernel would have to be thoroughly tested to insure no error was done.

@@

Several of the steps in the workflow can be automated: optimization and compiler
flags exploration, non regression testing, benchmarking... This automation can
be done by scripting by the user or by interfacing other dedicated tools. In
order to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution runtime.
\end{itemize}

@@

Usually computing kernels are hotspots of an HPC application, and they are most
of the time based on a loop nest. A lot of efforts are dedicated to their tuning
and the obtained result is often quite different from the original procedure.
Several transformations can be applied to such kernels. And those optimization
are often applied manually as compiler may fail to recognise the opportunity.

@@

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language four keywords were defined. They
are \emph{decl}, \emph{pr}, \emph{opn} and \emph{close}. As the language is
an EDSL, these four keywords are methods in the BOAST namespace.
%Usage of these keywords can be found on Figure~\ref{fig:keywords}.

@@

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of a length of $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to ruby
variables of the corresponding name.

@@

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
coma separated, similar to FORTRAN notation.

@@

Figure~\ref{fig:BOAST_control} shows some basic usages of the control structures.
The example a C macro or function that behaves similarly to the FORTRAN modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

The last control structure is Procedure. It is used to describe either procedure
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It is the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those array have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

@@

Similarly the target language must be changed with ease in order to compare
different alternatives. Two methods are dedicated to this task: \emph{set\_lang}
and \emph{get\_lang}. The target language can also be set through environment
variable before launching BOAST, allowing for easy command line scripting.

@@

Compilation of the generated kernels must also be very flexible. Indeed, HPC
applications may encounter platforms with very diverse compilation environment.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system exhibits similar behavior than common build system. Compilers
and their compile/build options can be specified. This can be done at several
places. Here is the list by increasing order of precedence: BOAST configuration
file, environment variables, and at kernel build time.

@@

The next logical step is to benchmark the built kernel. BOAST offers a simple
way to run a kernel that was successfully built while staying inside of BOAST.
A built BOAST \emph{CKernel} exposes a run method that accepts arguments corresponding
to the BOAST \emph{Procedure} used to instantiate the kernel. Arrays must be
instances of \emph{NArray} which are numerical arrays that use C array
underneath.

@@

Debugging applications running on GPU environments is well-recognized as a hard
and time-consuming activity. In complement with BOAST, we designed a trace-based
debugging tool that simplifies this porting operation. The tool relies on BOAST
support of multi-target code generation (Section~\ref{sec:multitarget}), used to
validate an application from one GPU-programming framework to another or between
differetn implementation using the same programming model. A case-study of the
port of a Cuda application to OpenCL is presented in
Section~\ref{subsec:specfem}.

The idea behind the tool is based on the assumption that the different
GPU ports of the code should do the very same operations, at least at
the logical the logical level (the APIs will have
\emph{implementation} differences, but they should offer nonetheless
same functionalities). The usage of BOAST framework sustains this
assumption, as both sets of kernels should generated from the same
source code.

@@

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of API
parameters are handles to opaque types, so, in order to provide
meaningful information, \code{GPUTrace} stores information about these
objects (handle value, buffer creation size and attributes, kernel
name and prototype, \etc{}).

Once two call traces have been generated by \code{GPUTrace}, it is
easy to compare them with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}.

@@

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimization can have an impact on the
performance of the kernel on the Mali architecture. But what of other
architectures? And are there other optimizations that can impact the
performance?

@@

The list of already identified optimizations are:
\begin{itemize}
\item vectorization,
\item intermediary data type,
\item number of pixels processed,
\item and synthesizing loads.
\end{itemize}
To create our generic implementation we decided to work at the component level
rather than the pixel level. This approach leads to more flexibility and
genericity when applying optimizations. We also decided to study the impact of
another parameter which is the number of component to process on the column
direction. This leads to being able to process tiles instead of only rows.

@@

Table~\ref{tbl:ARM} are the best results obtained by ARM on different image
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically beats the hand
optimized version. As far as the optimization options are concerned the result
are a disappointment: the same kernel configuration is the best for all image
sizes. This kernel is using \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} has no meaning here as \textit{x\_component\_number}
== \textit{vector\_length}. Those results also shows that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in 1 cycle without compromising
the bandwidth.

@@

But what if we run out benchmark on other architectures? Table~\ref{tbl:other}
the results obtained when running our BOAST implementation on other
architectures. The chosen architectures were an Intel i7-2760QM CPU (Sandy
Bridge architecture) that supports OpenCL 1.2 and an NVIDIA gtx680 GPU that
support OpenCL 1.1. We can see that the performance ratio between the different
architectures is rather stable across image sizes.

@@

In this Subsection we have shown the interest of BOAST in optimizing a well
known algorithm across different architectures. Chosen optimization combinations
are highly dependant on the targeted architecture. In the next Subsections we
will show that our methodology also applies to real applications.

@@

As a matter of fact, Daubechies wavelets exhibit a set of properties which make
them ideal for a precise and optimized DFT approach. In particular, their
systematicity allows to provide a reliable basis set for high-precision
results, whereas their locality (both in real and reciprocal space) is highly
desired to improve the efficiency and the flexibility of the treatment. Indeed,
a localized basis set allows to optimize the number of degrees of freedom for a
required accuracy~\cite{genovese2008daubechies}, which is highly desirable given
the complexity and inhomogeneity of the systems under investigation nowadays.

The application is mainly written in Fortran and currently includes about 360
thousands lines of Fortran code and 70 thousands lines of C language, which
accounts for more than 50 \% of the code base. It is a parallel application
based on the standards MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also
supports CUDA~\cite{cuda} and OpenCL~\cite{opencl}.  In the recent years this
code has been used for many scientific applications, and its development and
user consortium is continuously growing.  Massively parallel computations are
routinely executed with the BigDFT code, either in homogeneous or hybrid
architectures.  In 2009, the French Bull-Fourier was awarded for the
implementation of the hybrid version of BigDFT~\cite{genovese2009density}.

@@

The interest in having a robust and optimally tuned library in this scientific
field is therefore evident.  Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more aggressive inter procedural optimization can be performed.  This experience
therefore would be a first step toward the release of a tunable optimised
convolution library oriented to computational physics communities.

@@

Several kernels have already been implemented in BOAST for the convolution
library. Figure~\ref{fig:synthesis} show the performance of the wavelet
transform operation as a factor of the unrolling length of the outer loop, the
language used to implement it as well as the activation or not of the OpenMP
parallelization. Results are given as a speedup compared to the sequential hand
tuned implementation that can be found in BigDFT. These tests were run on the
Inte Xeon X5550 that was used to hand optimize the code.

@@

We can see that this function is better optimized using FORTRAN and small
unrolling factors. In the hand optimized version the unrolling factor was chosen
much too high (a factor of 12 was used). This factor might have been optimal at
the point the procedure was optimized (compiler version changed in the meantime) but
since then the environment changed. Other optimization have also been
incorporated in the BAOST sources, like the systematic inner loop unrolling, and
those could also help increase performance while limiting the interest of the
outer loop unrolling.

Nonetheless what is interesting from the physicist point of view is that the
generated source will give her better performance on a whole range of
architectures/compiler combination than that of the hand tuned code.

@@

This lack of \Specfem internal knowledge led us to be particularly
attentive to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

@@

As we had already validated BOAST-generated kernels (and trusted \Cuda
and \OCL versions to be semantically identical), we knew that the bugs
were now in the CPU side, and we had to find a way to understand where
\Specfem's \Cuda version of the code diverged from its \OCL
counterpart. To help us in that purpose, we had a strong assumption:
both versions of the code were supposed to perform exactly the same
operations, with the same ``logical'' parameters (the APIs have
\emph{implementation} differences, for instance \OCL has two memory
transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas \Cuda has only one, with a
direction parameter \code{cudaMemcpy(..., dir)}, but above that, it is
the same functionalities).

@@

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL runtime, as of
version 1.1.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.8 (from 7500 LOC to less than 4000, partly
because of code duplication, but by removing manually unrolled
loops). This is beneficial for \Specfem as it improves the readability
and maintainability of its source code.

@@

In the same mindset, we provided \Specfem testsuite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimization in kernel code and measure their impact, without
executing the whole application.

@@

  The most convenient way to obtain an application that can be auto-tuned on a
given platform is to base this application on a widely used computing library.
BLAS~\cite{dongarra1990set} and LAPACK~\cite{laug} are such libraries. These
libraries are either hand tuned for selected platforms or have auto-tuned
implementations. Atlas~\cite{whaley04} is an auto-tunes implementation of
BLAS/LAPACK. ATLAS authors defined the \emph{Automated Empirical Optimization of
Software} methodology that we implemented with BOAST. Their kernel generation is
done using macro-functions in C.

  Nonetheless many application formalisms cannot be reduced to standardized
library or border on what could be considered edge cases for those libraries and
not as optimized as more general cases. Orio~\cite{Hart2009:Orio} is an
auto-tuning framework that has an approach close of that of BOAST but they are
based on an annotated DSL describing loop transformation rather than a more
generic EDSL. Halide~\cite{ragan2013halide} is an auto-tuning framework
dedicated to image processing. It can also be used to describe other
operations on memory buffers. Those two frameworks propose automated search
space exploration to find the best version of a kernel.

@@

  Once auto-tuning technique are used, the parameter space explodes and the
systematic sampling rapidly becomes impossible to achieve. The cost of finding
the optimal kernel parameters and environment parameters (compiler flags, used
language, ...) is rapidly prohibitive. Dedicated frameworks have been developed
to address this problem. Adaptive Sampling Kit (ASK)~\cite{castro2013adaptive}
is one such tool. It reduces the number of sample needed by creating a model of
the performance and by minimizing the number of samples needed to find the
parameters of this model. Several models and sampling techniques are
implemented.

@@

Listing~\ref{lst:computing_kernel} shows how to create a Computing kernel
(\textit{CKernel}) and build it. Once a computing kernel is instancied the
output of BOAST will be redirected to the computing kernel source code. Line 4
sets the entry point of the computing kernel to the procedure we just defined.
By default compilation commands are not shown unless an error occurs. This
behavior can be change by switching to verbose mode.

When running the kernel all the arguments have to be specified. Running a
kernel returns a hash table containing information about the procedure
execution. In this simple case two information are returned, first the value
of the output parameter \textit{b} and second the time the kernel execution
took.

@@

Most computing kernels don't work on scalar values but rather on array of
data. Listing~\ref{lst:arrays} shows how to use arrays in computing kernels.
In this case we place ourselves in BOAST namespace to reduce the syntax
overhead. Variables \textit{a} and \textit{b} are one-dimensional arrays of
size \textit{n}. Arrays in BOAST start at index 1 unless specified otherwise.
For instance \lstinline!Dim(0,n-1)! would have created a dimension starting at
0. Array bounds can also be negative and several dimensions can be specified
to obtain muti-dimensional arrays.
For self contained procedures/kernels one can sue the shortcut written on
line~13 to create a CKernel object. As we are not specifying build options the
build command can also be ommited and will be automatically called when
running the kernel the first time. Lines~17 to 19 are used to check the result
of the kernel.

>>>>

\bibliographystyle{plain}
\begin{document}
\devnum{5.5}
\title{BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications}
\version{0.2}
\deadline{18}
\level{RE}
\nature{Report}
\coordinator{CNRS}
\contributors{Thierry Deutsch, Luigi Genovese, Jean-François Méhaut, Kevin Pouget, Brice Videau}
\reviewers{None}
\keywords{BOAST, Deliverable}

@@

\begin{changelog}
\change{0.1}{Initial version of the deliverable}
\change{0.2}{First version sent to reviewers}
\end{changelog}

@@

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in FORTRAN for historical and
performance reasons. Codes can be quite huge (several thousands lines of code)
with lots of functionalities. Nonetheless they are usually based on computing
kernels.  Computing kernels are resource intensive and well defined parts of a
program that usually work on precisely defined data. Those kernels are the time
consuming part of an HPC application. They are thus the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes some
of those developers only work a few monthes on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
FORTRAN may sometimes arrive later.

In Section~\ref{use_cases} we will present two HPC applications that we used
as use cases: SPECFEM3D and BigDFT. They are both based on computing
kernels and were selected as candidate applications in the Mont-Blanc project.

@@

The race to exascale is not going to simplify the environment. All of the above
architectures can be considered. Network architectures also can be very
diverse.  For instance European FP7 project DEEP considers using Accelerators
(XEON Phi) while the European FP7 project Mont-Blanc considers using low-power
embedded processor with integrated GPU.

@@

This is a good example of an algorithm that is simple in its formulation but
can be complex to optimize because many different optimizations are available
and can be combined together. The correct combination of optimizations will
depend on the targeted architecture.

@@

Listing~\ref{lst:laplace_c_filter} shows a simple implementation of the Laplace
filter in C99. For clarity, boundary conditions have been omitted.

@@

\paragraph{Vectorization}
The first proposed optimization is to compute 5 pixels instead of one using the
vectors available to the Mali architecture. To be efficient the vectors have to
be directly loaded from memory. Using vectors of 16 elements 15 useful components
are simultaneously loaded and can be used in computation. This optimization
yields speedups between 1.5 and 6 depending on the image size.

@@

\paragraph{More Optimizations} Several other optimizations can be attempted at
this point. For instance reducing or increasing the number of pixels each work
item is processing. They can yield improved performance in some cases, especially
when avoiding memory alignment problems. These optimizations will not be shown
here but can be found in~\cite{opencl_arm_training}.

@@

Nonetheless, the process is frustrating. The optimizations are never evaluated
independently from one another. Some were arbitrarily configured (the number of
pixels chosen for instance). Testing the different combinations of those
optimizations and the different parameters would prove very costly in developer
time and like all repetitive and tedious tasks, error prone. Thus every created
kernel would have to be thoroughly tested to ensure no error was done.

@@

Several steps in the workflow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking... This automation can be
done by scripting by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution runtime.
\end{itemize}

@@

Usually computing kernels are hotspots of an HPC application, and they are most
of the time based on a loop nest. A lot of efforts are dedicated to their tuning
and the obtained result is often quite different from the original procedure.
Several transformations can be applied to such kernels. And those optimizations
are often applied manually as compiler may fail to recognise the opportunity.

@@

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language four keywords were defined. They
are \emph{decl}, \emph{pr}, \emph{opn} and \emph{close}. As the language is an
EDSL, these four keywords are methods in the BOAST namespace. Sample usage of
these keywords will be found in the next Figures.
%Usage of these keywords can be found on Figure~\ref{fig:keywords}.

@@

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to ruby
variables of the corresponding name.

@@

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
comma separated, similar to FORTRAN notation.

@@

Figure~\ref{fig:BOAST_control} shows some basic usages of the control structures.
The example shows a C macro or function that behaves similarly to the FORTRAN modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

The last control structure is Procedure. It is used to describe either procedure
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It is the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those arrays have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

@@

Similarly the target language must be changed with ease in order to compare
different alternatives. Two methods are dedicated to this task: \emph{set\_lang}
and \emph{get\_lang}. The target language can also be set through an environment
variable before launching BOAST, allowing for easy command line scripting.

@@

Compilation of the generated kernels must also be very flexible. Indeed, HPC
applications may encounter platforms with very diverse compilation environments.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system exhibits similar behavior than common build systems. Compilers
and their compile/build options can be specified. This can be done at several
places. Here is the list by increasing order of precedence: BOAST configuration
file, environment variables, and at kernel build time.

@@

The next logical step is to benchmark the built kernel. BOAST offers a simple
way to run a kernel that was successfully built while staying inside of BOAST.
A built BOAST \emph{CKernel} exposes a run method that accepts arguments corresponding
to the BOAST \emph{Procedure} used to instantiate the kernel. Arrays must be
instances of \emph{NArray} which are numerical arrays that use C arrays
underneath.

@@

Debugging applications running on GPU environments is well-recognized as a hard
and time-consuming activity. In complement with BOAST, we designed a trace-based
debugging tool that simplifies this porting operation. The tool relies on BOAST
support of multi-target code generation (Section~\ref{sec:multitarget}), used to
validate an application from one GPU-programming framework to another or between
different implementations using the same programming model. A case-study of the
port of a Cuda application to OpenCL is presented in
Section~\ref{subsec:specfem}.

The idea behind the tool is based on the assumption that the different GPU
ports of the code should do the very same operations, at least at the logical
level (the APIs will have \emph{implementation} differences, but they should
offer nonetheless same functionalities). The usage of BOAST framework sustains
this assumption, as both sets of kernels should be generated from the same
source code.

@@

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of API
parameters are handles to opaque types, so, in order to provide
meaningful information, \code{GPUTrace} stores information about these
objects (handle value, buffer creation size and attributes, kernel
name and prototype, \etc{}). This is a custom trace format.

Once two call traces have been generated by \code{GPUTrace}, it is
easy to compare them with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}

@@

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimizations can have an impact on the
performance of the kernel on the Mali architecture. But what of other
architectures? And are there other optimizations that can impact the
performance?

@@

The list of already identified optimizations are:
\begin{itemize}
\item vectorization,
\item intermediary data type,
\item number of pixels processed,
\item and synthesizing loads.
\end{itemize}
To create our generic implementation we decided to work at the component level
rather than the pixel level. This approach leads to more flexibility and
genericity when applying optimizations. We also decided to study the impact of
another parameter which is the number of components to process on the column
direction. This leads to being able to process tiles instead of only rows.

@@

Table~\ref{tbl:ARM} are the best results obtained by ARM on different images
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically beats the hand
optimized version. As far as the optimization options are concerned the result
are a disappointment: the same kernel configuration is the best for all image
sizes. This kernel is using \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} has no meaning here as \textit{x\_component\_number}
== \textit{vector\_length}. Those results also shows that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in 1 cycle without
compromising the bandwidth. The results shown here are the best of 4 runs for
each configuration. As the test environment was well controlled kernel
durations were very stable.

@@

But what if we run our benchmark on other architectures? Table~\ref{tbl:other}
shows the results obtained when running our BOAST implementation on other
architectures. The chosen architectures were an Intel i7-2760QM CPU (Sandy
Bridge architecture) that supports OpenCL 1.2 and an NVIDIA gtx680 GPU that
support OpenCL 1.1. We can see that the performance ratio between the different
architectures is rather stable across image sizes.

@@

In this Subsection we have shown the interest of BOAST in optimizing a well
known algorithm across different architectures. Chosen optimization combinations
are highly dependent on the targeted architecture. In the next Subsections we
will show that our methodology also applies to real applications.

@@

As a matter of fact, Daubechies wavelets exhibit a set of properties which make
them ideal for a precise and optimized DFT approach. In particular, their
systematicity allows to provide a reliable basis set for high-precision
results, whereas their locality (both in real and reciprocal space) is highly
desired to improve the efficiency and the flexibility of the processing. Indeed,
a localized basis set allows to optimize the number of degrees of freedom for a
required accuracy~\cite{genovese2008daubechies}, which is highly desirable given
the complexity and inhomogeneity of the systems under investigation nowadays.

The application is mainly written in Fortran and currently includes about 360
thousands lines of Fortran code and 70 thousands lines of C language, which
accounts for more than 50 \% of the code base. It is a parallel application
based on the standards MPI~\cite{mpi} and OpenMP~\cite{openmp}.  It also
supports CUDA~\cite{cuda} and OpenCL~\cite{opencl}.  In the recent years this
code has been used for many scientific applications, and its development and
user consortium is continuously growing.  Massively parallel computations are
routinely executed with the BigDFT code, either in homogeneous or hybrid
architectures.  In 2009, the French Bull-Fourier award was attributed for the
implementation of the hybrid version of BigDFT~\cite{genovese2009density}.

@@

The interest in having a robust and optimally tuned library in this scientific
field is therefore evident. Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more aggressive inter procedural optimization can be performed. Indeed, BOAST
finds the optimal source code for a given kernel and compiler configuration but
one could imagine using binary optimizer to the compiled binaries. Those
optimizers could be coupled to BOAST and the output of the process would be the
final binary rather than the source code. This experience therefore would be a
first step toward the release of a tunable optimised convolution library
oriented to computational physics communities.

@@

Several kernels have already been implemented in BOAST for the convolution
library. Figure~\ref{fig:synthesis} shows the performance of the wavelet
transform operation as a factor of the unrolling length of the outer loop, the
language used to implement it as well as the activation or not of the OpenMP
parallelization. Results are given as a speedup compared to the sequential hand
tuned implementation that can be found in BigDFT. These tests were run on the
Intel Xeon X5550 that was used to hand optimize the code.

@@

We can see that this function is better optimized using FORTRAN and small
unrolling factors. In the hand optimized version the unrolling factor was chosen
much too high (a factor of 12 was used). This factor might have been optimal at
the time the procedure was optimized (compiler version changed in the meantime) but
since then the environment changed. Other optimizations have also been
incorporated in the BAOST sources, like the systematic inner loop unrolling, and
those could also help increase performance while limiting the interest of the
outer loop unrolling.

Nonetheless what is interesting from the physicist point of view is that the
generated source will give its better performance on a whole range of
architectures/compilers combinations than that of the hand tuned code.

@@

This lack of \Specfem internal knowledge led us to be particularly
careful to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

@@

As we had already validated BOAST-generated kernels (and trusted \Cuda and \OCL
versions to be semantically identical), we knew that the bugs were now in the
usage of the runtime, and we had to find a way to understand where \Specfem's
\Cuda version of the code diverged from its \OCL counterpart. To help us in
that purpose, we had a strong assumption: both versions of the code were
supposed to perform exactly the same operations, with the same ``logical''
parameters (the APIs have \emph{implementation} differences, for instance \OCL
has two memory transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas \Cuda has only one, with a direction
parameter \code{cudaMemcpy(..., dir)}, but above that, it is the same
functionalities).

@@

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL runtime, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in \OCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in \Cuda. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in \OCL.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.8 (from 7500 LOC to less than 4000, partly
because of code duplication, also removing manually unrolled
loops). This is beneficial for \Specfem as it improves the readability
and maintainability of its source code.

@@

In the same mindset, we provided \Specfem testsuite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimizations in kernel code and measure their impact, without
executing the whole application.

@@

  The most convenient way to obtain an application that can be auto-tuned on a
given platform is to base this application on a widely used computing library.
BLAS~\cite{dongarra1990set} and LAPACK~\cite{laug} are such libraries. These
libraries are either hand tuned for selected platforms or have auto-tuned
implementations. Atlas~\cite{whaley04} is an auto-tuned implementation of
BLAS/LAPACK. ATLAS authors defined the \emph{Automated Empirical Optimization of
Software} methodology that we implemented with BOAST. Their kernel generation is
done using macro-functions in C.

  Nonetheless many application formalisms cannot be reduced to standardized
library or border on what could be considered edge cases for those libraries and
not as optimized as more general cases. Orio~\cite{Hart2009:Orio} is an
auto-tuning framework that has an approach close of that of BOAST but they are
based on an annotated DSL describing loop transformation rather than a more
generic EDSL. Halide~\cite{ragan2013halide} is an auto-tuning framework
dedicated to image processing. It can also be used to describe other
operations on memory buffers. Those two frameworks propose automated search
space exploration to find the best version of a kernel.
LGen~\cite{spampinato2014basic} is a compiler that generates linear algebra
programs for small fixed size problems. Knowing the problem size it can fully
unroll and vectorize some loops, yielding better performance than state of the
art generic implementations.

@@

  Once auto-tuning techniques are used, the parameter space explodes and the
systematic sampling rapidly becomes impossible to achieve. The cost of finding
the optimal kernel parameters and environment parameters (compiler flags, used
language, ...) is rapidly prohibitive. Dedicated frameworks have been developed
to address this problem. Adaptive Sampling Kit (ASK)~\cite{castro2013adaptive}
is one such tool. It reduces the number of samples needed by creating a model of
the performance and by minimizing the number of samples needed to find the
parameters of this model. Several models and sampling techniques are
implemented.

@@

Listing~\ref{lst:computing_kernel} shows how to create a Computing kernel
(\textit{CKernel}) and build it. Once a computing kernel is instantiated the
output of BOAST will be redirected to the computing kernel source code. Line 4
sets the entry point of the computing kernel to the procedure we just defined.
By default compilation commands are not shown unless an error occurs. This
behavior can be changed by switching to verbose mode.

When running the kernel all the arguments have to be specified. Running a
kernel returns a hash table containing information about the procedure
execution. In this simple case two informations are returned, first the value
of the output parameter \textit{b} and second the time the kernel execution
took.

@@

Most computing kernels don't work on scalar values but rather on arrays of
data. Listing~\ref{lst:arrays} shows how to use arrays in computing kernels.
In this case we place ourselves in BOAST namespace to reduce the syntax
overhead. Variables \textit{a} and \textit{b} are one-dimensional arrays of
size \textit{n}. Arrays in BOAST start at index 1 unless specified otherwise.
For instance \lstinline!Dim(0,n-1)! would have created a dimension starting at
0. Array bounds can also be negative and several dimensions can be specified
to obtain muti-dimensional arrays.
For self contained procedures/kernels one can use the shortcut written on
line~13 to create a CKernel object. As we are not specifying build options the
build command can also be omitted and will be automatically called when
running the kernel the first time. Lines~17 to 19 are used to check the result
of the kernel.
